QPLIB: A Library of Quadratic Programming Instances

### 2016-08-04 QPLIB-1.0.tex
added a few citations, corrected a few typoes

In this paper we report the steps that have been done to collect a (hopefully) significant library of QP instances, filtering the large set of available (or specifically provided) instances in order to end up with a manageable set that still contains a meaningful sample of all possible QP types. A particularly thorny issue in this process is how to select instances that are ``interesting''. Our choica has been to take this to mean ``challenging for a significant set of solution methods''. Our filtering process has then been in part based on the idea that if a significant fraction of the solvers that can solve a QP instance do so in a ``short'' time, then the instance is not challenging enough to be included in the library. Yet, if very few (maybe one) solvers can solve it very efficiently by exploiting some specific structure, but most other approaches cannot, then the instance can still be deemed ``interesting''. Putting this approach in practice requires a nontrivial number of technical steps and decisions that are detailed in the paper. We hope that our work can provide useful guidelines for other interested researchers.

@@

Optimization problems with quadratic constraints and/or objective function (QP) have been the subject of a considerable amount of research over the last 60 years. At least some of the rationale for this interest is likely due to the fact that they are the ``least-nonlinear nonlinear problems''. Hence, in particular for the convex case, tools and techniques that have been honed during decades of research for Linear Programming (LP), typically with integrality constraints (MILP), can often be extended to the quadratic case with at least less effort than what would be required for tackling general Non Linear Programming (NLP) problems, without or with integrality constraints (MINLP). This has indeed happened over and over again with different algorithmic techniques, such as Interior Point methods, active-set methods (of which the simplex method is a prototypical example), enumeration methods, cut-generation techniques, reformulation techniques, and many others ({\bf citations?}). Similarly, nonconvex continuous QP are perhaps the ``simplest'' class of problems that require features like spatial enumeration techniques to be solved. Hence, they are both the natural basis for the development of general techniques for nonconvex NLP, and a very specific class so that specialized approaches can be developed \cite{Dur2010, Burer2012}.

On the other hand, (MI)QP is, in some sense, a considerably more expressive class than (MI)LP. Quadratic expressions are found, either naturally or after appropriate reformulations, in very many practical problems ({\bf citations?}). In general, any continuous function can be approximated with arbitrary accuracy (over a compact set) by a polynomial of arbitrary degree; in turn, every polynomial can be broken down to a system of quadratic expressions. Hence, (MI)QP is, in some sense, roughly as expressive as the whole of (MI)NLP. Of course this is, in principle, true for (MI)LP as well, but at the cost of much larger and much more complex formulations ({\bf citations?}). Hence, for many applications QP may represent the ``sweet spot'' between the effectiveness, but lower expressive power, of (MI)LP and the higher expressive power, but much higher computational cost, of (MI)NLP.

@@

The structure of this paper is the following. In \S \ref{sec:QPbasic} we review the basic notions about QP. In particular, \S \ref{subsec:notation} sets out the notation, \S \ref{sec:classification} proposes a---to the best of our knowledge, new---taxonomy of QP that helps us in discussing the (very) different classes of QPs, \S \ref{sec:algo} very briefly reviews the solution methods for QP upon which the solvers we have employed are based, and \S \ref{subsec:solver} describes the solvers. Then, \S \ref{sec:lib} describes the process used to obtain the library and its results; the software tools that we have used, and that are freely released together with the library, are discussed in \S \ref{subsec:tools}. Some conclusions are drawn in \S \ref{sec:conclusions}, after which in the Appendix a complete description of all the instances of the library is provided.

>>>>

In this paper we report the steps that have been done to collect a (hopefully) significant library of QP instances, filtering the large set of available (or specifically provided) instances in order to end up with a manageable set that still contains a meaningful sample of all possible QP types. A particularly thorny issue in this process is how to select instances that are ``interesting''. Our choice has been to take this to mean ``challenging for a significant set of solution methods''. Our filtering process has then been in part based on the idea that if a significant fraction of the solvers that can solve a QP instance do so in a ``short'' time, then the instance is not challenging enough to be included in the library. Yet, if very few (maybe one) solvers can solve it very efficiently by exploiting some specific structure, but most other approaches cannot, then the instance can still be deemed ``interesting''. Putting this approach in practice requires a nontrivial number of technical steps and decisions that are detailed in the paper. We hope that our work can provide useful guidelines for other interested researchers.

@@

Optimization problems with quadratic constraints and/or objective function (QP) have been the subject of a considerable amount of research over the last 60 years. At least some of the rationale for this interest is likely due to the fact that they are the ``least-nonlinear nonlinear problems''. Hence, in particular for the convex case, tools and techniques that have been honed during decades of research for Linear Programming (LP), typically with integrality constraints (MILP), can often be extended to the quadratic case with at least less effort than what would be required for tackling general Non Linear Programming (NLP) problems, without or with integrality constraints (MINLP). This has indeed happened over and over again with different algorithmic techniques, such as Interior Point methods, active-set methods (of which the simplex method is a prototypical example), enumeration methods, cut-generation techniques, reformulation techniques, and many others ({\bf citations?}). Similarly, nonconvex continuous QP are perhaps the ``simplest'' class of problems that require features like spatial enumeration techniques to be solved. Hence, they are both the natural basis for the development of general techniques for nonconvex NLP, and a very specific class so that specialized approaches can be developed \cite{Dur2010,Burer2012}.

On the other hand, (MI)QP is, in some sense, a considerably more expressive class than (MI)LP. Quadratic expressions are found, either naturally or after appropriate reformulations, in very many practical problems (\cite{Kochenberger2014}, {\bf citations?}). In general, any continuous function can be approximated with arbitrary accuracy (over a compact set) by a polynomial of arbitrary degree; in turn, every polynomial can be broken down to a system of quadratic expressions. Hence, (MI)QP is, in some sense, roughly as expressive as the whole of (MI)NLP. Of course this is, in principle, true for (MI)LP as well, but at the cost of much larger and much more complex formulations ({\bf citations?}). Hence, for many applications QP may represent the ``sweet spot'' between the effectiveness, but lower expressive power, of (MI)LP and the higher expressive power, but much higher computational cost, of (MI)NLP.

@@

The structure of this paper is the following. In \S~\ref{sec:QPbasic} we review the basic notions about QP. In particular, \S~\ref{subsec:notation} sets out the notation, \S~\ref{sec:classification} proposes a---to the best of our knowledge, new---taxonomy of QP that helps us in discussing the (very) different classes of QPs, \S~\ref{sec:algo} very briefly reviews the solution methods for QP upon which the solvers we have employed are based, and \S~\ref{subsec:solver} describes the solvers. Then, \S~\ref{sec:lib} describes the process used to obtain the library and its results; the software tools that we have used, and that are freely released together with the library, are discussed in \S~\ref{subsec:tools}. Some conclusions are drawn in \S~\ref{sec:conclusions}, after which in the Appendix a complete description of all the instances of the library is provided.

### 2016-08-04 QPLIB-2.2.tex
added a few citations, corrected a few typoes

Despite the apparent simplicity of the definition given in \S \ref{subsec:notation}, Quadratic Programming instances can be of several rather different ``types'' in practice, depending on the fine details of the data. In particular, many algorithmic approaches can only be applied to QP when the data of the problem has specific properties. A taxonomy of QP instances should therefore strive to identify the set of properties that an instance should have in order to apply the most relevant computational methods. However, the sheer number of different existing approaches, and the fact that new ones are proposed, makes it hard to provide a taxonomy that is both simple and covers all possible special cases. This is why, in this paper, we propose an approach that aims at finding a good balance between simplicity and coverage of the main families of computational methods.

@@

The ordering of the values in the previous lists is not irrelevant; in general, problems become ``harder'' when going from left to right. More specifically, for the \textit{F} and \textit{C} fields the order is that of \emph{strict} containment between problem classes: for instance, Linear objective functions are strictly contained in Diagonal convex quadratic ones (by just allowing the diagonal elements to be all-zero), which are strictly contained into general Convex quadratic ones (by allowing the off-diagonal elements to be all-zero), which in turn are strictly contained into general nonconvex Quadratic ones (by allowing any symmetric $Q^0$, hence possibly positive semidefinite ones as well). The only field for which the containment relationship is not a total order is \textit{V}, for which only the partial orderings
\[
 \mbox{C} \subset \mbox{M} \subset \mbox{G}
 \qquad,\qquad
 \mbox{B} \subset \mbox{M} \subset \mbox{G}
 \qquad,\qquad
 \mbox{B} \subset \mbox{I} \subset \mbox{G}
\]
hold. In the following discussion we will repeatedly exploit this by assuming that, unless otherwise mentioned, when a method can be applied to a given problem, it can be applied as well to all simpler problems where the value of each field is arbitrarily replaced with a value denoting a less-general class.

@@

  \item \emph{Nonconvex Binary Quadratic Programs} QB\{B, N, L\} obviously are $\mathcal{NP}$-hard. However, the special nature of binary variables combined with quadratic forms allows for quite specific techniques to be developed, among which is the reformulation of the problem as a LBL. Also, many well-known combinatorial problems can be naturally reformulated as problems of this class, and therefore a considerable number of results have been obtained by exploiting specific properties of the set of constraints ({\bf some citations? some specific problem to mention apart from Max-Cut?}).

@@

  \item \emph{Nonconvex Integer Quadratic Programs} QGQ is the most general, and therefore is the most difficul, class. Due to the lack of convexity even when integrality requirements are removed, solution methods must typically combine several algorithmic ideas, such as enumeration (distinguishing the role of integral variables from that of continuous ones involved into nonconvex terms) and techniques (e.g., outer approximation, SDP relaxation, \ldots) that allow to efficiently compute bounds. As in the convex case, QBQ, QMQ, and QIQ can benefit from more specific properties of the variables ({\bf citations?}).

>>>>

Despite the apparent simplicity of the definition given in \S~\ref{subsec:notation}, Quadratic Programming instances can be of several rather different ``types'' in practice, depending on the fine details of the data. In particular, many algorithmic approaches can only be applied to QP when the data of the problem has specific properties. A taxonomy of QP instances should therefore strive to identify the set of properties that an instance should have in order to apply the most relevant computational methods. However, the sheer number of different existing approaches, and the fact that new ones are proposed, makes it hard to provide a taxonomy that is both simple and covers all possible special cases. This is why, in this paper, we propose an approach that aims at finding a good balance between simplicity and coverage of the main families of computational methods.

@@

The ordering of the values in the previous lists is not irrelevant; in general, problems become ``harder'' when going from left to right. More specifically, for the \textit{O} and \textit{C} fields the order is that of \emph{strict} containment between problem classes: for instance, Linear objective functions are strictly contained in Diagonal convex quadratic ones (by just allowing the diagonal elements to be all-zero), which are strictly contained into general Convex quadratic ones (by allowing the off-diagonal elements to be all-zero), which in turn are strictly contained into general nonconvex Quadratic ones (by allowing any symmetric $Q^0$, hence possibly positive semidefinite ones as well). The only field for which the containment relationship is not a total order is \textit{V}, for which only the partial orderings
\[
 \mbox{C} \subset \mbox{M} \subset \mbox{G}
 \qquad,\qquad
 \mbox{B} \subset \mbox{M} \subset \mbox{G}
 \qquad,\qquad
 \mbox{B} \subset \mbox{I} \subset \mbox{G}
\]
hold. In the following discussion we will repeatedly exploit this by assuming that, unless otherwise mentioned, when a method can be applied to a given problem, it can be applied as well to all simpler problems where the value of each field is arbitrarily replaced with a value denoting a less-general class.

@@

  \item \emph{Nonconvex Binary Quadratic Programs} QB\{B, N, L\} obviously are $\mathcal{NP}$-hard. However, the special nature of binary variables combined with quadratic forms allows for quite specific techniques to be developed, among which is the reformulation of the problem as a LBL. Also, many well-known combinatorial problems can be naturally reformulated as problems of this class, and therefore a considerable number of results have been obtained by exploiting specific properties of the set of constraints ( \cite{Rendl2008}, {\bf some citations? some specific problem to mention apart from Max-Cut?}).

@@

  \item \emph{Nonconvex Integer Quadratic Programs} QGQ is the most general, and therefore is the most difficul, class. Due to the lack of convexity even when integrality requirements are removed, solution methods must typically combine several algorithmic ideas, such as enumeration (distinguishing the role of integral variables from that of continuous ones involved into nonconvex terms) and techniques (e.g., outer approximation, SDP relaxation, \ldots) that allow to efficiently compute bounds. As in the convex case, QBQ, QMQ, and QIQ can benefit from more specific properties of the variables (\cite{Buchheim2013,Dong2016}, {\bf citations?}).

### 2016-08-04 QPLIB-2.3.tex
added a few citations, corrected a few typoes

The solvers in Table \ref{t:solvers} which implement incomplete methods are {\sc ConOpt}, {\sc IpOpt}, {\sc Minos}, {\sc SnOpt} and {\sc KNitro}. The former four solvers implement local search algorithms for continuous nonconvex NLPs (a problem class containing QCQ). Note that traditionally {\sc KNitro} used to be a local (nonconvex) NLP solver. All of these solvers implement essentially three types of local search methods for NLP:
%
\begin{itemize}
 \item active set methods ({\sc ConOpt}, {\sc Minos}, {\sc SnOpt}) \cite{Dost09};
 %
 \item interior point methods ({\sc IpOpt}, {\sc KNitro}) \cite{Wright97};
 %
 \item projected gradient methods \cite{CalaMore87:mp, ChenGui13:coap}.
\end{itemize}
%
Active set and interior point methods have been defined for *C* but also for general (continuous) NLPs. In fact, all of the solvers named above target this larger problem class.

@@

The solvers in Table \ref{t:solvers} BB type methods are:
%
\begin{itemize}
 \item {\sc Antigone}, {\sc Baron}, {\sc Couenne}, {\sc Lindo}, {\sc LindoGlobal}, {\sc SCIP}, {\sc lgo\_bb}, which implement complete BB algorithms for QGQ;
 %
\item {\sc CPLEX}, which implements a complete BB algorithm for QGL;
%
 \item {\sc Knitro\_bb}, {\sc BonMin}, {\sc sBB}, {\sc Xpress}, {\sc Gurobi}, and {\sc Mosek} which implement complete BB algorithms for CGC.}
\end{itemize}
%
We remark that the latter category can be used as incomplete solvers for QGQ. We also remark that CPLEX can currently only target problems with linear constraints when the objective function is nonconvex \cite{bliek}.

>>>>

The solvers in Table \ref{t:solvers} which implement incomplete methods are {\sc ConOpt}, {\sc IpOpt}, {\sc Minos}, {\sc SnOpt} and {\sc KNitro}. The former four solvers implement local search algorithms for continuous nonconvex NLPs (a problem class containing QCQ). Note that traditionally {\sc KNitro} used to be a local (nonconvex) NLP solver. All of these solvers implement essentially three types of local search methods for NLP:
%
\begin{itemize}
 \item active set methods ({\sc ConOpt}, {\sc Minos}, {\sc SnOpt}) \cite{Dost09};
 %
 \item interior point methods ({\sc IpOpt}, {\sc KNitro}) \cite{Wright97};
 %
 \item projected gradient methods \cite{CalaMore87:mp,ChenGui13:coap}.
\end{itemize}
%
Active set and interior point methods have been defined for *C* but also for general (continuous) NLPs. In fact, all of the solvers named above target this larger problem class.

@@

The solvers in Table \ref{t:solvers} BB type methods are:
%
\begin{itemize}
 \item {\sc Antigone}, {\sc Baron}, {\sc Couenne}, {\sc Lindo}, {\sc LindoGlobal}, {\sc SCIP}, {\sc lgo\_bb}, which implement complete BB algorithms for QGQ;
 %
\item {\sc CPLEX}, which implements a complete BB algorithm for QGL;
%
 \item {\sc Knitro\_bb}, {\sc BonMin}, {\sc sBB}, {\sc Xpress}, {\sc Gurobi}, and {\sc Mosek} which implement complete BB algorithms for CGC.
\end{itemize}
%
We remark that the latter category can be used as incomplete solvers for QGQ. We also remark that CPLEX can currently only target problems with linear constraints when the objective function is nonconvex \cite{bliek}.
